---
title: "Prediction basics"
format: html
editor: visual
---

#Prediction basics


### Data

```{r}
library("easypackages")
packages("tidyverse","fpp3", "tsibble", "feasts","fable", "patchwork")
```
Data will work with
```{r}
global_economy |> 
  filter(Country == "Sweden") |> 
  autoplot(GDP) +
  ggtitle("Sweden PIB") + ylab("$US billions")
```
### Model definition

```{r}
TSLM()

```

```{r}
fit <- global_economy |> 
  model(trendModel = TSLM(GDP ~ trend()))
  
```
```{r}
fit
```

## Predicting

```{r}
fcst <- fit |> forecast(h="3 years")
fcst
```
```{r}
fcst |> 
  filter(Country == "Sweden") |> 
  autoplot()
```
```{r}
fcst |> 
  filter(Country == "Sweden") |> 
  autoplot(global_economy)
```
####Simple prediction methods

```{r}
bricks <- aus_production |> 
  filter_index("1970" ~ "2004")
bricks
```
```{r}
bricks |> 
  autoplot(Bricks)
```
#### Mean prediction
```{r}
bricks |> model(MEAN(Bricks))
```


```{r}
train <- aus_production |> filter_index("1992 Q1" ~ "2006 Q4")

beerFit <- train |> 
  model(
    Mean = MEAN(Beer),
    `Naïve`= NAIVE(Beer),
    `Seasonal naïve` = SNAIVE(Beer),
    Drift = RW(Beer~ drift()) #Randow walk includes a linear component
  )

#Predict
beerPredictions <- beerFit |> forecast(h = 14)

#Plot forecast


beerPredictions %>%
  autoplot(filter_index(aus_production, "1992 Q1" ~ .), level = NULL) +
  ggtitle("Forecasts for quarterly beer production") +
  xlab("Year") + ylab("Megalitres") +
  guides(colour=guide_legend(title="Forecast")) +
  geom_vline(xintercept = as.Date("2007-01-01"), color = "firebrick",
             linetype = "dashed") +
  annotate("label", x = c(as.Date("2003-01-01"),as.Date("2009-01-01")),
           y = 550, label = c("Train set", "Test set"),
           color = c("black","blue"))
```

Plotting just predictions

```{r}
autoplot(beerPredictions)
```
Season Naive prediction
```{r}
beerPredictions %>%
  filter(.model == "Seasonal naïve") %>% 
  autoplot(filter_index(aus_production, "1992 Q1" ~ .)) +
  ggtitle("Seasonal naïve forecast for quarterly beer production") +
  xlab("Year") + ylab("Megalitres") +
  guides(colour=guide_legend(title="Forecast")) +
  geom_vline(xintercept = as.Date("2007-01-01"), color = "firebrick",
             linetype = "dashed") +
  annotate("label", x = c(as.Date("2003-01-01"),as.Date("2009-01-01")),
           y = 550, label = c("Train set", "Test set"),
           color = c("black","blue"))
```
#### Another exaple

```{r}
gafa_stock |> distinct(Symbol)
```
```{r}
#Re-index based on trading days
googleStock <- gafa_stock |> 
                filter(Symbol == "GOOG") %>%
                mutate(day = row_number()) |> 
                update_tsibble(index = day, regular = TRUE)

#Filter the year of interest
google_2015 <- googleStock |> filter(year(Date) == 2015)

#Fit the model
googleFit <- google_2015 |> 
  model(
      Mean    = MEAN(Close),
    `Naïve` = NAIVE(Close),
    Drift   = NAIVE(Close ~ drift()),
    SNAIVE  = SNAIVE(Close)
  )
```
```{r}
googleFit
```
```{r}
#Produce forecast for the first 19 trading days in 2016
googleForecast <- googleFit |> forecast(h=19)

googleForecast
```
A better way to use a tsiblle to determine forecast
```{r}
#timeframe to predict
googleJan2016 <- googleStock |> 
  filter(yearmonth(Date) == yearmonth("2016 Jan"))

#Asking to predict values from the timeframe defined previously
googleForecast <- googleFit |> forecast(googleJan2016)

#plot the forecast
googleForecast %>%
  autoplot(google_2015, level = NULL) +
    autolayer(googleJan2016, Close, color='black') +
    ggtitle("Google stock (daily ending 31 Dec 2015)") +
    xlab("Day") + ylab("Closing Price (US$)") +
    guides(colour=guide_legend(title="Forecast"))
```
Comparing Drift versus NAIVE
```{r}
p1 <- googleForecast %>%
  filter(.model == "Drift") %>% 
  autoplot(google_2015) +
    autolayer(googleJan2016, Close, color='black') +
    ggtitle("Google stock (daily ending 31 Dec 2015)") +
    xlab("Day") + ylab("Closing Price (US$)") +
    guides(colour=guide_legend(title="Forecast"))

p2 <- googleForecast %>%
  filter(.model == "Naïve") %>% 
  autoplot(google_2015) +
    autolayer(googleJan2016, Close, color='black') +
    ggtitle("Google stock (daily ending 31 Dec 2015)") +
    xlab("Day") + ylab("Closing Price (US$)") +
    guides(colour=guide_legend(title="Forecast"))

p1 / p2
```
## Fitted values and residuals

Each observation in a timeseries could be predicted using previous historic values, this is known as fitted values

Model residuals is the different between fitted values(predictions) and real values

```{r}
#We can use augment to get fitted values and residuals
augment(beerFit)
```
#### Residual 

A good prediction will give you residuals with the following characteristics

- Not correlated
- Residuals mean is 0

Useful but not fundamental carachteristics

- Constant variance between them
- Normally distributed

Box Cox could help with this

```{r}
google_2015 |> 
  autoplot(Close) +
  xlab("Day") + ylab("Closing Price (US$)") +
  ggtitle("Google Stock in 2015")
```
We'll use NAIVE method
```{r}
aug <- google_2015 |> 
  model(NAIVE = NAIVE(Close)) |> 
  augment()

aug
```
```{r}
#Pulling the average of the residuals without taking nulls into consideration

aug |> pull(.resid) |> mean(na.rm = TRUE)
```
```{r}
p1 <- aug %>% autoplot(.resid) + xlab("Día") + ylab("") +
  ggtitle("NAIVE residuals")

p2 <- google_2015 |> 
  autoplot(Close) +
  xlab("Day") + ylab("Closing Price (US$)") +
  ggtitle("Google Stock in 2015")

p2/p1
```
We have the biggest residuals where there were drastic changes
```{r}
aug |> ggplot(aes(x= .resid)) +
  geom_histogram() +
  ggtitle("Residuals histogram")
```
## Testing if residuals are correlated
```{r}
aug %>% ACF(.resid)
```
```{r}
aug %>% ACF(.resid) %>% autoplot() + ggtitle("ACF of residuals")
```
#### Using NAIVE model

```{r}
#function to get info about residuals
google_2015 |> 
  model(NAIVE(Close)) |> 
  gg_tsresiduals()
```
```{r}
aus_production %>% 
  filter_index("1992" ~ .) %>%
  model(SNAIVE(Beer)) %>% 
  gg_tsresiduals()
```
```{r}
aus_production %>% 
  filter_index("1992" ~ .) %>%
  gg_tsdisplay(Beer)
```
Autocorrelation function shows that residuals are not correlated. It seems that NAIVE method is capturing all important information

#### Using mean model
```{r}
google_2015 |> 
  model(MEAN(Close)) |> 
  augment() |> 
  pull(.resid) |> 
  mean(na.rm = TRUE)
```
```{r}
#residual diagnostic

google_2015 |> 
  model(MEAN(Close)) |> 
  gg_tsresiduals() + 
  ggtitle("Residuals diagnostic for mean method")
```
-Correlations are big
-Residuals follow a pattern
-Histogram is not normally distributed

#### Methods to evaluate correlation
```{r}
#Evaluating NAIVE model
#We want the box_pierce value to be small
aug |> features(.resid, box_pierce, lag=10, dof=0)
```

```{r}
#We want the Ljung test to be big
aug %>% features(.resid, ljung_box, lag=10, dof=0)
```
This show us that residuals aren't correlated


Now, let's test the same but using mean model

```{r}
google_2015 |> 
  model(MEAN(Close)) |> 
  augment() |> 
  features(.resid, box_pierce,lag=10, dof=0 )
```

```{r}
google_2015 |> 
  model(MEAN(Close)) |> 
  augment() |> 
  features(.resid, ljung_box,lag=10, dof=0 )
```

In both tests, value is big so residuals are not distinguishable from noise


## Prediction Intervals

The bigger the interval, the less precise our model.
Precise prediction by themself is not useful, so we add the foreast interval

There are various types of intervals:
  - One step interval: When forecast is just one value, standar deviation from forecast is the same for the residuals
  - multi-step interval: When forecast steps increase, the forecast interval tends to get bigger
  
```{r}
google_2015 |> 
  model(NAIVE(Close)) |> 
  forecast(h=10) |> 
  hilo()
```
```{r}
google_2015 |> 
  model(NAIVE(Close)) |> 
  forecast(h=10) |> 
  autoplot(google_2015)
```
#### Bootstrapping to generate different scenarios by resampling data

```{r}
fit <- google_2015 |> 
  model(NAIVE(Close))

sim <- fit |> generate(h = 30, times=5, bootstrap =TRUE, seed = 123) #5 possible scenarios for next 30 trading days
sim
```

```{r}
google_2015 %>%
  ggplot(aes(x = day)) +
  geom_line(aes(y = Close), size = 1) +
  geom_line(aes(y = .sim, colour = as.factor(.rep)), data = sim, size = 1) +
  ggtitle("Google closing stock price") +
  guides(col = FALSE)
```
```{r}
fc1 <- fit |> forecast(h=30, bootstrap = TRUE) |> autoplot(google_2015) + ggtitle("Google closing stock price")


fc2 <- fit |> forecast(h=30) |> autoplot(google_2015) + ggtitle("Google closing stock price") #Not bootstrapped

fc1 / fc2

```
  